---
title: "Owen cumulative functions"
author: "Stéphane Laurent"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Owen cumulative functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(OwenQ)
knitr::opts_chunk$set(collapse=TRUE)
```


## The Owen distribution

Let $Z \sim {\cal N}(0,1)$ and $X \sim \chi^2_\nu$ be two independent random variables.
For real numbers $\delta_1$ and $\delta_2$, define the two random variables
$$
T_1 = \frac{Z+\delta_1}{\sqrt{\dfrac{X}{\nu}}}
\quad \text{and} \quad\;
T_2 = \frac{Z+\delta_2}{\sqrt{\dfrac{X}{\nu}}}.
$$

Both $T_1$ and $T_2$ follow a non-central Student distribution. 
The number of degrees of freedom is $\nu$ for each of them, and their respective non-centrality parameters are $\delta_1$ and $\delta_2$ respectively. 

Owen (1965) studied the distribution of the pair $(T_1, T_2)$.

The four Owen cumulative functions are
$$
\begin{align}
O_1(\nu, t_1, t_2, \delta_1, \delta_2) & = \Pr(T_1 \leq t_1, T_2 \leq t_2), \\
O_2(\nu, t_1, t_2, \delta_1, \delta_2) & = \Pr(T_1 \leq t_1, T_2 \geq t_2), \\
O_3(\nu, t_1, t_2, \delta_1, \delta_2) & = \Pr(T_1 \geq t_1, T_2 \geq t_2), \\
O_4(\nu, t_1, t_2, \delta_1, \delta_2) & = \Pr(T_1 \geq t_1, T_2 \leq t_2).
\end{align}
$$

Owen provided a way to evaluate these functions *when $\nu$ is an integer number*. Owen's algorithms are implemented in the `OwenQ` package. 

For $\delta_1 > \delta_2$, these four functions are implemented in the `OwenQ` package under the names `powen1`, `powen2`, `powen3` and `powen4`. 
For general values of $\delta_1$ and $\delta_2$, they are implemented under the names `psbt1`, `psbt2`, `psbt3` and `psbt4` 


## Non-central Student distribution 

Owen (1965) also provided an algorithm to evaluate the cumulative distribution function of an univariate non-central Student distribution with an integer number of degrees of freedom. 
This evaluation is performed by the function `ptOwen` of the `OwenQ` package.

```{r}
ptOwen(q=1, nu=3, delta=2)
pt(q=1, df=3, ncp=2)
```

It is known that the `pt` function is not reliable when the non-centrality parameter `ncp` is large. 
Below we compare the values given by `ptOwen` and `pt` and the value given by Wolfram (`N[CDF[NoncentralStudentTDistribution[4,70],80],20]`):

```{r}
print(pt(q=80, df=4, ncp=70), digits=20)
print(ptOwen(q=80, nu=4, delta=70), digits=20)
print(0.54742763380700947685, digits=20) # Wolfram value
```

## Limitations

When `q`$=$`delta`, the value of `ptOwen(q, nu, delta)` should go to `0.5` as `nu` increases to infinity. The examples below show the failure of this expectation when `nu` is too large.

```{r}
ptOwen(q=50, nu=3500, delta=50)
ptOwen(q=50, nu=3600, delta=50)
ptOwen(q=50, nu=3650, delta=50)
ptOwen(q=50, nu=3660, delta=50)
ptOwen(q=50, nu=3670, delta=50)
ptOwen(q=50, nu=3680, delta=50)
```


## Application to equivalence testing 

Assume a statistical model given by a sample 
$y_i \sim_{\text{iid}} {\cal N}(\mu, \sigma^2)$ for $i=1, \ldots, n$. 
We are interested in the alternative hypothesis 
$H_1\colon\{\Delta_1 \leq \mu \leq \Delta_2\}$. 

Consider the $100(1-2\alpha)\%$-confidence interval about $\mu$:
$$
\bigl[\bar y - t^\ast_{n-1}(\alpha)\frac{\hat\sigma}{\sqrt{n}}, \, 
\bar y + t^\ast_{n-1}(\alpha)\frac{\hat\sigma}{\sqrt{n}} \bigr].
$$

The $H_1$ hypothesis is accepted if this interval falls into the interval 
$[\Delta_1, \Delta_2]$.

This can be written as follows:
$$
T_1 := 
\frac{\bar y - \Delta_1}{\hat\sigma/\sqrt{n}} 
\geq t^\ast_{n-1}(\alpha)
\quad \text{and} \quad 
T_2 := 
\frac{\bar y - \Delta_2}{\hat\sigma/\sqrt{n}} 
\leq - t^\ast_{n-1}(\alpha).
$$

Observe that 
$$
T_1 = \frac{z - \delta_1}{\dfrac{\sqrt{n-1}\hat\sigma/\sigma}{\sqrt{n-1}}}
$$
where
$$
z = \frac{\sqrt{n}\bar y}{\sigma} - \mu \sim {\cal N}(0,1)
\quad \text{and} \quad 
\delta_1 = \frac{\mu - \Delta_1}{\frac{\sigma}{\sqrt{n}}}.
$$

By reasoning in the same way for $T_2$, we find that $(T_1, T_2)$ follows the 
Owen distribution with degrees of freedom $\nu = n-1$, and non-centrality 
parameters $\delta_1$ given above and 
$$
\delta_2 = \frac{\mu - \Delta_2}{\frac{\sigma}{\sqrt{n}}}
$$

Therefore the power of the test - *i.e.* the probability to accept $H_1$ - is given by 
$$
O_4(n-1, t^\ast_{n-1}(\alpha), -t^\ast_{n-1}(\alpha), \delta_1, \delta_2).
$$


### Inconclusive equivalence test

The result of the equivalence test is said to be *inconclusive* when only one of the bounds of the confidence interval falls into the interval $[\Delta_1, \Delta_2]$. 

The probability to get an inconclusive result can be obtained with the `OwenQ` package. We show that with the help of simulations below.

```{r}
Delta1 <- -2; Delta2 <- 2 
mu <- 1; sigma <- 6; n <- 30L
alpha <- 0.05
nsims <- 1e5L
equivalence <- inconclusive <- numeric(nsims)
for (i in 1L:nsims) {
  y <- rnorm(n, mu, sigma)
  CI <- t.test(x = y, conf.level = 1-2*alpha)$conf.int
  equivalence[i] <- (CI[1] > Delta1) && (CI[2] < Delta2)
  inconclusive[i] <- ((CI[1] < Delta1) && (CI[2] > Delta1)) ||
    ((CI[1] < Delta2) && (CI[2] > Delta2))
}
```

```{r}
dof <- n-1
q <- qt(1-alpha, dof)
se <- sqrt(1/n)*sigma
delta1 <- (mu-Delta1)/se; delta2 <- (mu-Delta2)/se
# probability to get equivalence
mean(equivalence)
powen4(dof, q, -q, delta1, delta2)
# probability to get inconclusive
mean(inconclusive)
ptOwen(q, dof, delta2) - ptOwen(-q, dof, delta1) - powen4(dof, q, -q, delta1, delta2)
```


## The Owen $T$-function

The `OwenQ` package also provides an implementation of the Owen $T$-function, 
under the name `OwenT`. 
This is a port of the function `owens_t` of the C++ `boost` library.
